---
title: "DOT project summary"
output: 
   html_document:
    toc: true
    code_folding: hide
---
---

```{r setup, include=FALSE}
source("setup.R")


# ABSTRACT NOTES
# https://docs.google.com/document/d/1ZfwtacaTX3zldbmpV0rx868fiasGo-2dF3ZMkSWYk6s/edit?usp=sharing


  
  # 2016 OPM fedscope text files
  # https://www.dropbox.com/s/xe74anownix98bn/FederalEmployees.csv?dl=0
```

```{r global.options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      cache = F, 
                      fig.width=8.5, 
                      fig.align = 'center',
                      fig.path='Figs/',
                      warning=FALSE, message=FALSE)
```

# Abstract

Administrative rulemaking is often conceptualized as a straightforward translation of legislative delegations into agency rules, where bureaucrats adopt their most preferred policy given their constraints, often speeding up or slowing down the policy process to do so. The rulemaking process is indeed highly formulaic, requiring publication of a draft, external review, and public comment before a final rule is published. However, much of the politics of rulemaking may play out in the less formal and transparent process before a draft policy is published. In this stage, different priorities and constraints may drive decisions about which draft policies are published and when.

Theories of policymaking suggest that the politics of rulemaking may differ significantly in the pre-proposal stage. Yet, this part of the policy process has largely eluded quantitative study due to a lack of systematic data. This paper begins to fill this gap. We introduce a new dataset based on thousands of monthly internal project status reports by the U.S. Department of Transportation that allow us to track a rulemaking effort from its origin date through its finalization.  Additionally, we develop a number of novel measures of prioritization and constraints that we are able to measure before and after a draft rule is made public,  including the degree of agency resource constraints, internal agency prioritization, changes in sub-agency political appointees, congressional and presidential attention, and the threat of litigation and sanction by courts.

Using these new data and measures, we model the drivers of prioritization and constraint as a multi-stage policy process and identify different political forces at work in each stage. Specifically, we use a multi-stage hazard model to test factors that may affect the timing of rules being proposed or finalized thus identify predictors of prioritization and delay. Previous research suggests that that greater internal agency resources—in terms of personnel and budgets—will speed up rulemaking during the largely opaque pre-proposal stage, while attention from external actors, including Congress and the president, will speed up rulemaking during the more transparent notice and comment period. Our data and measures allow new tests of these theories. We show that resources, complexity, and salience have different effects on the speed of rule-writing across different stages.

Our findings hold important implications for our understanding of transparency and bureaucratic policymaking within a representative democracy.




---

# To-do:

### Data

- Find a method to identify the year of each rule's authorizing statute

- Add OIRA report dates 

- Infer missing withdrawal dates from the date last observed in stage withdrawal

- Add subagency administrators and secretary data

- OPM staff resources data for years other than 2016 

### Model

- Properly code observations for previous and future stages

- Add covariates

---



# Data

```{r data}
load(here("data/DOT-monthly.Rdata"))
dotMonthly %<>% filter(!is.na(initiated)) # select rule-month observations with a dot report

load(here("data/DOT-perRule-perStage.Rdata"))
dotStage %<>% filter(!is.na(initiated)) # select rule-stage observations with a dot report

load(here("data/DOT-perRule.Rdata"))
dotRIN %<>% filter(!is.na(initiated)) # select rule observations with a dot report

dotMonthly %<>% group_by(RIN) %>% 
  mutate(delay = paste(na.omit(color), collapse = ",")) %>%
  mutate(delayedever = grepl("Red|Yellow", delay)) 
```

[Google spreadsheet of variables](https://docs.google.com/spreadsheets/d/1CptLu6Xz7K14KF8PyELsiy-siYZHpi5jjmS9hMyO1xM/edit?usp=sharing)

### Summary Stats
```{r summarystats}
# HOW MANY WITH REASON (NOTE NEED TO CORRECT WHYDELAY CONCAT)
# dotRIN$whydelay1 <- gsub("N/A",NA,dotRIN$whydelay1)
# sum(!is.na(dotRIN$whydelay1))
sum(!is.na(dotRIN$whydelay))  %>%
write("stats/RINsWithReasonsForDelay.tex")

# HOW MANY DELAYED
filter(dotRIN, color %in% c("Red", "Yellow")) %>%
  n_distinct(.$RIN) %>%
  write("stats/RINsDelayed.txt")

# HOW MANY NOT DELAYED WHEN PUBLISHED
filter(dotRIN, color %in% c("Black", "Green")) %>%
  n_distinct(.$RIN) %>%
  write("stats/RINsNotDelayed.tex")

# HOW MANY AT EACH STAGE 
n_distinct(dotRIN$RIN)   %>% write("stats/uniqueRINs.tex")

sum(!is.na(dotRIN$ANPRMpublished))   %>% write("stats/ANPRMs.tex")

sum(!is.na(dotRIN$ANPRM_RECIEVED))  %>% write("stats/ANPRMs-recieved-by-OMB.tex")

sum(!is.na(dotRIN$ANPRMtoOST))  %>% write("stats/ANPRMs-to-OST.tex")

sum(!is.na(dotRIN$NPRMpublished))   %>% write("stats/NPRMs.tex")

sum(!is.na(dotRIN$NPRMtoOST))  %>% write("stats/NPRMs-to-OST.tex")

sum(!is.na(dotRIN$NPRM_RECIEVED))  %>% write("stats/NPRMs-recieved-by-OMB.tex")

sum(!is.na(dotRIN$SNPRMpublished))   %>% write("stats/SNPRMs.tex")

sum(!is.na(dotRIN$SNPRM_RECIEVED))  %>% write("stats/SNPRMs-recieved-by-OMB.tex")

sum(!is.na(dotRIN$IFRpublished))   %>% write("stats/IFRs.tex")

sum(!is.na(dotRIN$IFR_RECIEVED))  %>% write("stats/IFRs-recieved-by-OMB.tex")

sum(!is.na(dotRIN$FinalRulePublished))   %>% write("stats/FinaldotRIN.tex")

sum(!is.na(dotRIN$FINAL_RECIEVED))  %>% write("stats/FinaldotRIN-recieved-by-OMB.tex")

sum(!is.na(dotRIN$WithdrawalPublished))  %>% write("stats/Withdrawals.tex")



#############################################################
# days-until-EVENT 
#############################################################
dotRIN %>% 
  filter(!is.na(DaysUntilANPRM)) %>% 
  summarise(mean(DaysUntilANPRM)) %>%
  first()   %>% write("stats/days-until-ANPRM.tex")

dotRIN %>% 
  filter(!is.na(DaysUntilNPRM)) %>% 
  summarise(mean(DaysUntilNPRM)) %>%
  first()   %>% write("stats/days-until-NPRM.tex")

dotRIN %>% 
  filter(!is.na(DaysUntilFinalRule)) %>% 
  summarise(mean(DaysUntilFinalRule)) %>%
  first()   %>% write("stats/days-until-final.tex")


dotRIN %>% 
  filter(!is.na(DaysUntilOSTANPRM)) %>% 
  summarise(mean(DaysUntilOSTANPRM)) %>%
  first()   %>% write("stats/days-until-ANPRM-to-OST.tex")

dotRIN %>% 
  filter(!is.na(DaysUntilOSTNPRM)) %>% 
  summarise(mean(DaysUntilOSTNPRM)) %>%
  first()   %>% write("stats/days-until-NPRM-to-OST.tex")

dotRIN %>% 
  filter(!is.na(DaysUntilOSTFinalRule)) %>% 
  summarise(mean(DaysUntilOSTFinalRule)) %>%
  first()   %>% write("stats/days-until-Final-to-OST.tex")

dotRIN %>% 
  filter(!is.na(DaysUntilWithdrawal)) %>% 
  summarise(mean(DaysUntilWithdrawal)) %>%
  first()  %>% write("stats/days-until-Withdrawal.tex")

# how many initiated post-2008
dotRIN %>% 
  filter(initiated > as.Date("2008-01-01")) %>% 
  #filter(!is.na(FinalRulePublished)) %>%  
  n_distinct(.$RIN) %>%
  first() %>% write("stats/initiatedSince2008.tex")

  # unique(dotRIN$prompt)
```
  


We have initiation dates for `r length(unique(dotRIN$RIN))` unique RINs. Of these, `r sum(dotRIN$color == "Red", na.rm = T)` were marked 'delayed,'`r sum(dotRIN$color == "Yellow", na.rm = T)` were marked 'potentially delayed,' and `r sum(dotRIN$color == "Green", na.rm = T)` were marked 'on time' at the time they were last observed. However, DOT increasingly marks projects as 'not on a schedule' (`r sum(dotRIN$color == "Black", na.rm = T)` RINs were last observed in the “Black” category). `r dim(filter(dotMonthly, delayedever == T, !is.na(DOTdate)) %>% select(RIN) %>% distinct())[1]` projects were delayed at some point and `r sum(!is.na(dotRIN$whydelay))` of these give a reason for delay.

`r sum(dotRIN$initiated > as.Date("2008-01-01"), na.rm = T)` RINs were initiated after 2008. `r sum(grepl("Final Rule", dotRIN$STAGE), na.rm = T)` have Final Rules Published, `r sum(dotStage$STAGE == "Withdrawal", na.rm =T)` have been withdrawn, and `r sum(dotStage$STAGE == "Termination", na.rm =T)` were terminated.



This table shows the average number of months after initiation at which certian milestones are reached:

| *Average months until:*   | to Secretary| to ORIA | Published | *N* |
| ------------------ | ------------- | ---------------- | -------------------- | --------- |
|ANPRM              | `r round(mean(dotStage$DaysUntilOSTANPRM, na.rm = T)/30)` |`r round(mean(dotStage$DaysUntilOMBANPRM, na.rm = T)/30)` | `r round(mean(dotStage$DaysUntilANPRM, na.rm = T)/30)`| n = `r sum(dotStage$STAGE == "Prerule", na.rm =T)` |
|NPRM               |`r round(mean(dotStage$DaysUntilOSTNPRM, na.rm = T)/30)` |`r round(mean(dotStage$DaysUntilOMBNPRM, na.rm = T)/30)` | `r round(mean(dotStage$DaysUntilNPRM, na.rm = T)/30)`| n = `r sum(dotStage$STAGE == "Proposed Rule", na.rm =T)` |
|SNPRM              |`r round(mean(dotStage$DaysUntilOSTSNPRM, na.rm = T)/30)` |`r round(mean(dotStage$DaysUntilOMBSNPRM, na.rm = T)/30)` | `r round(mean(dotStage$DaysUntilSNPRM, na.rm = T)/30)`| n = `r sum(dotStage$STAGE == "SNPRM", na.rm = T)` |
|Interim Final Rule  |`r round(mean(dotStage$DaysUntilOSTIFR, na.rm = T)/30)` |`r round(mean(dotStage$DaysUntilOMBIFR, na.rm = T)/30)` | `r round(mean(dotStage$DaysUntilIFR, na.rm = T)/30)`| n = `r sum(dotStage$STAGE == "Interim Final Rule", na.rm =T)`|
|Final Rule         |`r round(mean(dotStage$DaysUntilOSTFinalRule, na.rm = T)/30)` |`r round(mean(dotStage$DaysUntilOMBFinalRule, na.rm = T)/30)` | `r round(mean(dotStage$DaysUntilFinalRule, na.rm = T)/30)`| n = `r sum(grepl("Final Rule", dotRIN$STAGE), na.rm = T)`|
|Withdrawal         |`r round(mean(dotStage$DaysUntilOSTWithdrawal, na.rm = T)/30)` |`r round(mean(dotStage$DaysUntilOMBWithdrawal, na.rm = T)/30)` | `r round(mean(dotStage$DaysUntilWithdrawal, na.rm = T)/30)`| n = `r sum(dotStage$STAGE == "Withdrawal", na.rm =T)`|
|Termination        |||| n = `r sum(dotStage$STAGE == "Termination", na.rm =T)`|

\Sexpr{knit_child('stats/days-until-ANPRM-to-OST.tex')}



---

Adding DOT Internal Status Reports to OIRA and Unified Agenda Reports gives a richer picture of rulemaking. Data by RIN, date, and source:
```{r timeline, fig.height = 17}
# If project is ongoing, define enddate as date last observed (for plotting active project status)
dotMonthly %<>% mutate(end = if_else( is.na(enddate), reportdate, enddate) )

# Consolidate like stages for plotting:
dotMonthly$STAGE %<>% {gsub("Terminat.*", "Withdrawal",.)}
dotMonthly$STAGE %<>% {gsub(".2$", "", .)}
dotMonthly$STAGE %<>% {gsub(".3$", "", .)}
dotMonthly$STAGE %<>% {gsub(".4$", "", .)}
dotMonthly$STAGE %<>% {gsub("Notice", "Proposed Rule", .)}
dotMonthly$STAGE %<>% {gsub("Other", "Undetermined", .)}
dotMonthly$STAGE %<>% {gsub("Prerule", "Proposed Rule", .)}
dotMonthly$STAGE %<>% {gsub("Interim Final Rule", "Final Rule", .)}
dotMonthly$STAGE %<>% {gsub("SNPRM", "Proposed Rule", .)}

dotMonthly %>% # filter(reportdate > as.Date("2008-01-01")) %>%
ggplot(aes(x = RIN)) + 
  geom_linerange(aes(ymin=initiated, ymax=end, color="DOT Reports"), size=.1) +
  geom_point(aes(y = DOTdate, color = "DOT Reports", shape = STAGE), size = .5) +
  geom_point(aes(y = ANPRMpublished, color = "DOT Reports"), shape = 2, size = .5) +
  geom_point(aes(y = NPRMpublished, color = "DOT Reports"), shape = 2, size = .5) +
  geom_point(aes(y = SNPRMpublished, color = "DOT Reports"), shape = 2, size = .5) +
  geom_point(aes(y = FinalRulePublished, color = "DOT Reports"), shape = 1, size = .5) +
  geom_point(aes(y=IFRpublished, color = "DOT Reports"), shape = 1, size = .5) +
  geom_point(aes(y=WithdrawalPublished), color="red", shape = 4, size = 1) +
  geom_point(aes(y=Terminated), color="red4", shape = 4, size = 1) +
  geom_point(aes(y = DATE_RECEIVED, color = "OIRA Reports", shape = STAGE), size = .5) +
  geom_point(aes(y = UnifiedAgendaDate, color = "Unified Agenda Reports", shape = STAGE), size = .5) +
  facet_grid(acronym~., scales = "free_y", space = "free_y")+
  coord_flip() +
  scale_y_date(lim = c(as.Date("2000-01-01"), as.Date("2019-01-01")),
             breaks=date_breaks(width = "1 year"), 
             labels = date_format("%y"),
             minor_breaks=NULL) + 
  ggtitle('Significant DOT Rulemaking Projects Active Between 2008 and 2017') + 
  labs( x = "RIN", y = "Date", color = "", shape = "")+
  theme(axis.text.y = element_text(size = 3,
                       angle = 0,
                       hjust = 1,
                       vjust = 1),
        axis.ticks = element_blank(),
        strip.text.y = element_text(angle = 0),
        strip.background = element_blank(),
        legend.position = "top") 
```

Notice how the DOT (esp. FTA and FHA) initiated a large number of projects shortly after Obama's reelection, probably waiting for assurance that they would be able to complete their work.  


---

```{r stages-hist}
# stage histogram 
ggplot(dotMonthly %>% filter(STAGE %in% c("Withdrawal","Proposed Rule", "Other", "Final Rule", "Interim Final Rule"))) +
  stat_count(aes(DOTdate)) +
  facet_grid(STAGE~., scales = "free_y") +
  labs(title = "Active per Month by Stage", x = "")
  #scale_y_date(breaks=date_breaks(width = "1 year"), labels = date_format("%y"), minor_breaks=NULL)
```

```{r hist-stages-published, include=FALSE}
# publication histogram 

  ggplot(dotRIN %>% filter(STAGE %in% c("Withdrawal","Proposed Rule", "Final Rule", "Interim Final Rule")), 
         aes(publicationactual, na.rm = T)) +
    geom_histogram(na.rm = T, bins = length(unique(dotRIN$year)))  + 
    facet_grid(STAGE~., scales = "free_y") +
  labs(title = "Published by Stage", x = "")
  #scale_y_date(breaks=date_breaks(width = "1 year"), labels = date_format("%y"), minor_breaks=NULL)
```

```{r completion-time-by-promt}
# Consolidate like stages for plotting:
dotRIN$STAGE %<>% {gsub("Terminat.*", "Withdrawal",.)}
dotRIN$STAGE %<>% {gsub(".2$", "", .)}
dotRIN$STAGE %<>% {gsub(".3$", "", .)}
dotRIN$STAGE %<>% {gsub(".4$", "", .)}
dotRIN$STAGE %<>% {gsub("Notice", "Proposed Rule", .)}
dotRIN$STAGE %<>% {gsub("Other", "Undetermined", .)}
dotRIN$STAGE %<>% {gsub("Prerule", "Proposed Rule", .)}
dotRIN$STAGE %<>% {gsub("Interim Final Rule", "Final Rule", .)}
dotRIN$STAGE %<>% {gsub("SNPRM", "Proposed Rule", .)}

  dotRIN %>% filter(STAGE %in% c("Final Rule", "Withdrawal")) %>%  mutate(end = if_else( is.na(enddate) & DOTdate != "2018-09-01", reportdate, enddate) ) %>% #filter(prompt %in% c("Secretarial/Head of Operating Administration Decision","2011 Retrospective Regulatory Review" , "International Agreement", Settlement Agreement") ) %>% 
  ggplot() +  
  geom_point(aes(x = initiated, y = end, color = prompt, shape = STAGE)) + 
  scale_y_date(limits = c(as.Date("2008-01-01"),as.Date("2019-01-01")),
               breaks=date_breaks(width = "1 year"), 
             labels = date_format("%Y") )+

    #geom_smooth(aes(x= if_else(initiated > as.Date("2008-01-01"), initiated, NA), y= enddate), method = "lm" ) +
  labs(title = "Completed Rulemaking Projects by Prompting Reason and Outcome", x = "Initiated", y = "Published", color = "Prompt to Initiate Project", shape = "")
```






---



# Models

We are interested in two kinds of outcomes: (1) what happens? and(2) how long does it take?

Given these questions and the timeseries nature of data (montly status reports with changing covariate values), we use duration models. Because we are interested in how policies may move through multiple stages of the policymaking process, we use a multi-state model. Doing so allows us to simultaniously model how different factors may affect a policy at different stages, while also modeling a common fraility across rules. We could model each stage seperately, but this would overlook correlation within rules across stages. 

Two interpretable statistics address each of our main questions. 

First, we estimate "transition probabilities" (i.e. the probablity that a policy will advance to each possible future state at a given time) using the from Aalen-Johansen estimator, also known as the cumulative incidence function (the function estimating the number of observations that have transitioned to and from each state over time). 

Second, we estimate the probability that a given transition will occur in a given time window using the Kaplan-Meier estimator. Covariates may increase or decrease the liklihood that a given transition will occur, thus increaseing or decreasing the estimated rate, $\lambda$, of advancing from a given stage to a given stage in the policymaking process, and thus the Kaplan-Meier estimate that a given transition will occur after a given date. 

To capture common elements of each rule, we use a penalised likelihood a random effects (frailty) model, with frailties for each rulemaking project (each Regultory Identification Number). 


## Core Model
First, we model the most common stages that rules go through with no covariates, only the dates of transition in and out of the pre-NPRM and NPRM stages. We use the DOT data with one observation per rule per stage.

Formatting data for multi-stage duration modeling requires identifying the dates of transitions to and from each stage. To do this we must define and index a set of transitions of interest. The data are formatted as one observation per rule per transition. The key variable is the days since the rulemaking project was initiated at which that transition occured. 

```{r clean-for-2stage-4trans-model}
# select vars for minimal model 
d <- select(dotStage, DOTdate, RIN, STAGE, initiated, DaysUntilNPRM, DaysUntilFinalRule, DaysUntilWithdrawal) %>%
  # select minimal stages
  filter(!is.na(initiated), !is.na(DOTdate), STAGE %in% c("Proposed Rule", "Final Rule", "Withdrawal")) %>% 
  distinct() %>% 
  arrange(DOTdate) %>% arrange(RIN)

# Days until event 
# FIXME
# note, bellow subs last observed date as published date where missing; not always correct
d %<>% mutate(DaysUntilNPRM = ifelse(is.na(DaysUntilNPRM) & STAGE == "Proposed Rule", DOTdate - initiated, DaysUntilNPRM))
d %<>% mutate(DaysUntilFinalRule = ifelse(is.na(DaysUntilFinalRule) & STAGE == "Final Rule", DOTdate - initiated, DaysUntilFinalRule))
d %<>% mutate(DaysUntilWithdrawal = ifelse(is.na(DaysUntilWithdrawal) & STAGE == "Withdrawal", DOTdate - initiated, DaysUntilWithdrawal))


# identify previous stages in the data (need to use previous stage var to supplement this)
d %<>% group_by(RIN) %>%
  mutate(stages = paste(STAGE, collapse = ";")) %>% ungroup() 

# head(d$stages)
# unique(d$stages)

# FIXME

# NEED TO INDEX TRANSITIONS 
# e.g. trans Initiated to NPRM = 1, Straight to final = 2, NPRM withdrawal = 3,  NPRM published to FINAL published = 4

d$trans <- NA
# to proposed rule = trans 1 
d$trans[grepl("^Proposed Rule", d$stages) & d$STAGE == "Proposed Rule"] <- 1

# direct to final = trans 2 
d$trans[d$stages =="Final Rule"] <- 2

# withdrawal = trans 3
d$trans[d$stages == "Proposed Rule;Withdrawal" & d$STAGE == "Withdrawal"] <- 3

# proposed to final = trans 4 
d$trans[grepl("Proposed Rule;Final Rule", d$stages) & d$STAGE == "Final Rule"] <- 4

# Transition Matrix 
tmat <- transMat(list(c(2, 4), #transitions from prerule
                      c(3,4), 		 		  #transitions from nprm
                      c(), 		 		  #transitions from withdrawal
                      c()),        # trans from final 
                 names = c("preNPRM", "NPRM", "Withdrawal", "Final"))


# USE INDEX TO ID ENTRY AND EXIT DATES  
# prerule stage: entry = initiated, exit = NPRM published, status = 1 if DOT report date = NPRM published date

# obs entering at date initiated
d %<>% mutate(entry = ifelse(trans %in% c(1,2), 0, NA))

# obs entering at NPRM date 
d %<>% mutate(entry = ifelse(trans %in% c(3,4), DaysUntilNPRM, entry))

# obs exiting at Final Rule published date
d %<>% mutate(exit = ifelse(trans %in% c(2,4), DaysUntilFinalRule, NA))

# obs exiting at Withdrawal published date
d %<>% mutate(exit = ifelse(trans %in% c(3), DaysUntilWithdrawal, exit))

# obs exiting at NPRM published date
d %<>% mutate(exit = ifelse(trans %in% c(1), DaysUntilNPRM, exit))

# NOTE THERE ARE SOME ERRORS THAT MAY NEED TO BE INVESTIGATED IF THEY PERSIST AFTER PROPER CODING OF TRANSITIONS 
d %<>% filter(exit > entry) # FIXME

# ALL OBS ARE TRANSITIONS, HOWEVER, WE CAN INCLUDE NON TRANSITION OBS (i.e. months where no transition happened) -- DOES IT HELP? Perhaps only with covariats that have values for those non-transition months?
d$status <- 1
```

In estimation, a multi-stage cox hazard model is a cox hazard model stratified by transition.

Click on "Code" to see model -> 
```{r model-2stage-4trans}
# MODEL WITH NO COVARIATES 
mod.1 <- coxph(Surv(entry, exit, status) ~ strata(trans), data = d, method = "breslow")

# # INSPECT 
# summary(mod.1)
# class(mod.1$xlevels)

# # Extract fitted values
fit <- msfit(mod.1, trans=tmat, variance=TRUE)

# CAUTION NOTE: in one run, Withdrawals got dropped, making 4 (NPRM to Final) the 3rd factor. Not cool.
f <- fit$Haz

# Name stage for trans 1 and 2
f %<>% mutate(Stage = ifelse(trans %in% c(1,2), " Pre NPRM Stage", "Post NPRM Stage") )

# Name transitions
f$trans %<>% {gsub(1, "Inititiated to NPRM", .)}
f$trans %<>% {gsub(2, "Direct to Final", .)}
f$trans %<>% {gsub(3, "NPRM to Withdrawal", .)} 
f$trans %<>% {gsub(4, "NPRM to Final", .)} 

# Name variables
names(f) <- c("Days", "Hazard", "Transition", "Stage")
```

Plot hazard rates (the probability that a rule at a given stage hits a each possible transition after so many days at that stage) for simple 2-stage, 4-transition model:
```{r model-1}
# plot
ggplot(f) +
  geom_line(aes(x = Days, y = Hazard, color = Transition)) + 
  facet_grid(. ~ Stage) + 
  theme_minimal()
``` 

However, there are more possible transitions. Here are the most common patterns: 

```{r most-common-transitions, include=F}
# identify previous stages in the data (need to use previous stage var to supplement this)
# lots of obs have "completed actions as the stage; hoping we can do better; this is not what we wanted to get from UA/ORIA data"
dotStage %>%
  distinct() %>% 
  filter(!is.na(initiated)) %>% 
  arrange(DOTdate)%>% 
  filter(STAGE != "Completed Actions") %>% 
  group_by(RIN) %>%
  mutate(stages = paste(STAGE, collapse = "; ")) %>% ungroup() %>% 
  mutate(stages = paste("Initiated;", stages)) %>% 
  group_by(stages) %>% summarise(N = n()) %>% select(N, stages) %>%
  filter(N>4) %>% 
  arrange(-N) %>% 
  kable() #%>% kable_styling()
```

And without "Undetermined" status:
```{r most-common-transitions-named, include=FALSE}
# identify previous stages in the data (need to use previous stage var to supplement this)
# lots of obs have "completed actions as the stage; hoping we can do better; this is not what we wanted to get from UA/ORIA data"
dotStage %>%
  distinct() %>% 
  filter(!is.na(initiated)) %>% 
  arrange(DOTdate)%>% 
  filter(!STAGE %in% c("Undetermined","Completed Actions") )%>% 
  group_by(RIN) %>%
  mutate(stages = paste(STAGE, collapse = "; ")) %>% ungroup() %>% 
  mutate(stages = paste("Initiated;", stages)) %>% 
  group_by(stages) %>% summarise(N = n()) %>% select(N, stages) %>%
  filter(N>4) %>% 
  arrange(-N) %>% 
  kable() #%>% kable_styling()
```

## Adding covariates

```{r model-2}
# move to merge 
dotRIN %<>% mutate(Same_President = ifelse(initiated>as.Date(2009-01-01) & enddate<as.Date(2017-01-01), "Yes", "No")) 
dotRIN %<>% mutate(Same_President = ifelse(is.na(Same_President), "No", Same_President))
dotRIN %<>% mutate(MAJOR = ifelse(MAJOR == "Undetermined", "No", MAJOR))

############################
# Here we take event data with rows indexed by ID and columns for relevent dates and covariates.
# The method takes date variables, melts the data to identify the order in which various transitions occur, and, finally, models transitions in and out of each stage as a multi-state survival model. 
# THis is a slightly more generic approach than the mstate package vinettes as it does not require one to specifiy which transitions are possible or actually present in the data. Instead, the transition matrix is derived from the data.
# It also attempts to keep data in a tidy format (rather than special class objects). 

# First we identify the order in which events occured in order to identify different types of transitions. 

# STEP 1: IDENTIFY TRANSITIONS AND DATES 

# These data start as one observation per ID (in this case, "RIN")
# Every observation must have a start date (here the date the rulemaking project was "initiated").
d <- filter(dotRIN, !is.na(initiated)) 

# We split out transitions observed for each ID (We will merge back in with covariates later.)
# Select the events we want to model:
d %<>% select(RIN, initiated, NPRMpublished, FinalRulePublished, WithdrawalPublished)

# Melt data to be one observation per event
d %<>% 
  melt(id = "RIN", na.rm = T, value.name = "transition_date") %>% 
  distinct()

# Convert dates to strings for now to make things easier (specifically, the super-useful mutate(ifelse()) method). 
d$transition_date %<>% as.character()
# Group by RIN and identify sequence
d %<>% 
  group_by(RIN) %>% 
  arrange(transition_date) %>% 
  rename(event = variable) %>%
  mutate(transitions = paste(event, collapse = " to ")) 

# Parse transitions from sequence (could be made more general?)
# FIXME
d %<>% 
  mutate(trans1 = str_match(transitions, "([A-z]*? to [A-z]*){1}")[1] ) %>% 
  mutate(trans2 = str_match(transitions, "([A-z]*? to [A-z]*){2}")[1] ) %>%
  mutate(trans2 = str_replace(trans2, "(.*? to ){1}", "") ) %>%
  mutate(trans3 = str_match(transitions, "([A-z]*? to [A-z]*){3}")[1] ) %>%
  mutate(trans3 = str_replace(trans3, "(.*? to ){2}", "") )

d %<>%  select(-transitions) 

d %<>% melt(id = c("RIN", "event", "transition_date"), na.rm = T, value.name = "transition") %>% 
  rename(transition_count = variable)

d %<>% 
  group_by(RIN, transition) %>% 
  mutate(exit_date = ifelse(event == gsub(".* to ", "", transition),
                            transition_date, NA)) %>% 
  mutate(exit_date = paste(unique(na.omit(exit_date)), collapse = ";")) %>%
  mutate(entry_date = ifelse(event == gsub(" to .*", "", transition),
                            transition_date, NA)) %>% 
  mutate(entry_date = paste(unique(na.omit(entry_date)), collapse = ";"))

# Now that we have identified transitions and the entry and exit dates, we no longer need the `event` and `transition_date` variables 
# However, we will need the event names later, so we save them
events <- as.character(unique(d$event))
d %<>% select(-event, -transition_date) %>% distinct()



# STEP 2: MERGE BACK WITH DATA

d %<>% left_join(dotRIN) %>% ungroup() %>% arrange(RIN) 

# Count days until event with `difftime()`
d %<>% 
  mutate(exit = difftime(exit_date, initiated, units="days") ) %>% 
  mutate(entry = difftime(entry_date, initiated, units="days") ) 

# Days until initiated is obviously 0, but difftime has odd rounding, so let us fix that:
d$entry[grepl("^init", d$transition)] <- 0
  
##########################################################
# INDEX TRANSITIONS 
d$transition %<>% as.factor()

d %<>% mutate(trans = factor(transition) ) %>% 
  mutate(trans = factor(trans, levels = rev(levels(trans))))

# See the levels before converting them to numbers
trans <- tlevels <- levels(d$trans)
# events

# Possible transitions from state i
for(i in 1:length(events)){
  trans <- gsub(events[i], i, trans)
}

# Make a matrix of from and to
trans <- str_split(trans, " to ", simplify = TRUE)

# Use the matrix to populate the list for the transMat()
from <- list()
for(i in 1:length(events)){
  from[[i]] <- as.numeric(trans[which(trans[,1]==i),2])
}

# Transition Matrix
tmat <- transMat(from, 
                 names = events)


# NOTE THERE ARE SOME ERRORS (at least one left) THAT MAY NEED TO BE INVESTIGATED IF THEY PERSIST AFTER PROPER CODING OF TRANSITIONS 
d %<>% filter(exit > entry) # FIXME

# ALL OBS ARE TRANSITIONS, HOWEVER, WE CAN INCLUDE NON TRANSITION OBS (i.e. months where no transition happened) -- DOES IT HELP? Perhaps only with covariats that have values for those non-transition months?
d$status <- 1
# ```

# In estimation, the multi-stage cox hazard model is a cox hazard model stratified by transition. 
library(broom)

fit <- survfit(Surv(entry, exit, status) ~ MAJOR + strata(trans), data = d)
# summary(fit)
# tidy(fit) 
tidyfit <- fit %>% 
  tidy() %>%
  separate(strata, c("Major", "strata"), ", ") %>% 
  #filter(estimate > 0) %>%
  filter(time<3650) %>%
  mutate(strata = gsub(".*=|published|Published| *$","", strata) ) %>%
  mutate(from = factor(paste("from",gsub(".*=| to .*","", strata)) ) ) %>% 
  mutate(from = factor(from, levels = rev(levels(from)))) %>%
  mutate(to = gsub(".*=|.* to ","to ", strata) ) # %>% 

ggplot(tidyfit, aes(time/365, estimate)) + 
  geom_line(aes(color = Major)) +
  scale_x_continuous(breaks = seq(0,10, by = 1)) +
  geom_point(aes(color = Major), shape = "+") + 
  geom_ribbon(aes(fill = Major, ymin = conf.low, ymax = conf.high), alpha=.25) + 
  facet_grid(to~ from, scales = "free_x") + 
  labs(x = "Years" , y = "KM Probability that transition will take longer than t")

```
---



## Covariates

#### Covariates: 

[Google spreadsheet of variables](https://docs.google.com/spreadsheets/d/1CptLu6Xz7K14KF8PyELsiy-siYZHpi5jjmS9hMyO1xM/edit?usp=sharing)

Some are measured at the rule level, others at the agency level, and others for all rules at a given point in time. We may need to cluster standard errors by rule or agency depending on the covariates.

##### Resources 
Resources measures:

- sub-agency budget? - can we get the budget for rulemaking activities? Is this different than estimating SES? 

- "lack of resources" or "lack of staffing" claimed [0,1], subset of delayed rules

- number of SES / # of rules per year (ideally at sub-agency level if we have SES data for that)

- secretary prompt [0,1] - how many missing? 

- Non-acting subagency head [0,1]

- Same president as initiation date [0,1]

- Same secretary as initiation date [0,1]

- Same subagency head as initiation date [0,1]

##### Complexity  

Complexity measurements:

- "coordination necessary" claimed [0,1], subset of delayed rules

- “unanticipated issues/impacts requiring further analysis”*“awaiting development of additional data” claimed [0,1], subset of delayed rules

- the score of the abstract on Benoit's political sophistication/complexity

- second comment period [1,0] (though this may also be a measure of external, interest group constraint) 

- the number of comments? (to me this is more about salience and contentiousness than complexity)

- MORE MEASURES? 

##### Contentiousness: 
 White House and Secretary alignment measures:
 
- "other, higher priorities" claimed [0,1] (is it right to interpret this way? Priority could also be internal...which would make it, what? Not a resource per se)

- initiated under same president [0,1] (same party president? we have a few started under Clinton, finished under Obama) 

- went to Secretary twice [0,1]

- days between going to secretary going to ORIA (average if multiple times)

- went to ORIA twice [0,1]

- Did ORIA make changes [0,1]

- days between going to ORIA and publication (count days from the first or last time at ORIA? )

Congress alignment measures:

- positive-negative sentiment words in subcommittee budget report

- positive-negative sentiment words in the subcommittee hearing

- legislative deadline exists [0,1]

- legislative deadline past [0,1]

- prompting action is legislation [0,1] (speeds up rulemaking?)

- legislation is pending [0,1] (slows rulemaking)

- Reg flex required [0,1] slowing? 

- unified congress 

Litigation threat:

- court deadline exists [0,1]

- court deadline past [0,1]

- number of lawsuits against the subagency in the past x years

- avg number of lawsuits against the agency per year of administration (this is perhaps better suited for a model that is not just DOT)

- litigation threatened in comments? [0,1]

## Congressional attention/opposition: 

Below are examples of positive and negative sentiment about the FAA from the 2018 appropriations subcommittee report. Much it is very neutral, but it is fair to say that this actually reflects neutral attitudes. Sentiment analysis allows us to tally up the positive and negative words for each subagency. Other scholars have measured the length of the section on each subagency, but this mostly just reflects the size and diversity of its budget.

#### Examples:

POSITIVE: 

The Committee is encouraged 
by the additional measures the FAA is taking to enhance 
outreach to communities affected by new flightpaths. The 
Committee recommendation includes an additional $2,000,000 to 
support the FAA's ongoing efforts to address community noise 
concerns.
The Committee supports 
research that is being conducted through the FAA's Center of 
Excellence for Alternative Jet Fuel and Environment, the 
Aviation Sustainability Center (ASCENT) on the impact of 
aviation noise on both sleep and cardiovascular health.
The Committee commends the FAA Office 
of Commercial Space Transportation's efforts to promote private 
sector lunar exploration and development and encourages the FAA 
to explicitly define non-interference and to enhance its 
payload review process to provide companies planning private 
sector lunar development with the security and predictability 
necessary to support substantial investments. 

NEGATIVE:

The Committee is concerned by the FAA's decision to 
extend by five years its aging Mission Support Network, a 
decision which carries both technology and cost implications. 



# Notes

#### DOT - Unified Agenda rule stage crosswalk

List of unique rule stages and how we code each (Note /2 means the second such document for the same RIN):

 [1] "Final Rule"  = Final Rule                         
 [2] "Undetermined" = Undetermined                        
 [3] "NPRM" = Proposed Rule                                
 [4] "SNPRM" = SNPRM (sometimes folded into Proposed Rule)
 [5] "Interim Final Rule/4" = Interim Final Rule                
 [6] "Interim Final Rule" = Interim Final Rule                  
 [7] "Withdrawal" = Withdrawal                          
 [8] "Other" = Undetermined                               
 [9] "ANPRM" = Prerule                               
[10] "Final Rule/2" = Final Rule/2                        
[11] "NPRM/2"= Proposed Rule                               
[12] "SNPRM/2"= SNPRM                             
[13] "Notice" = Notice                              
[14] "Supplemental Notice of Intent" = Prerule       
[15] "Reconsideration of Final Rule" = Final Rule      
[16] "Disposition of Comments" = Other            
[17] "Request for Comments" = Other                
[18] "Final Rule with Request for Comments" = Final Rule
[19] "Termination"  = Termination (sometimes folded into Withdrawal)



---


